---
title: "A factor tutorial in R"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = "figures/",
                      echo = TRUE, warning = FALSE, message = FALSE)
```

---

Click [here](http://htmlpreview.github.io/?https://github.com/nickgood/a_little_factor_tutorial/blob/master/factor_tutorial.html) to view the marked-up html file.

# Packages

```{r}
library(forcats)
library(dplyr)
library(readr)
```

---

# Making Factors

Factors are a categorical vectors of variables with an order mapped onto them. By default `R` will apply alphabetical ordering to a factor.

We can create factors using the `factor()` function:

```{r factor}
treatments <- factor(c("lpg", "three_stone", "gasifier",
                       "fan_rocket", "rocket_elbow", "control"))
treatments
```

We can check use `parse_factor()` to see if the levels we are creating match a vector of factors. This is useful when checking for anomolies in a dataset for example:

```{r parse, warning = TRUE}
x <- c("control", "lpg", "lpg", "gasifier", "control",
       "fan_rocket", "rocket elbow", "rocket_elbow")

levels <- c("lpg", "three_stone", "gasifier",
            "fan_rocket", "rocket_elbow", "control")

parse_factor(x, levels)
```

Note how the variable in the vector `x` is coerced to `<NA>` and a warning message is given. Note also, there is a level slot for `three_stone_fire` in the new object, even though it doen't appear in the data vector `x`.

---

# Summarizing Factors

We'll often encounter a column of class `factor` in a dataframe e.g.: 

```{r}
stoves <- data_frame(type = factor(sample(levels, 500, replace = TRUE)))

head(stoves, 4)
```

We can use the `levels()` functions to return a character vector containing the distinct levels the a factor contains. We can then interrogate the vector returned by `levels()`:

```{r}
levels(stoves$type)

class(levels(stoves$type))

length(levels(stoves$type))
```

We can use `group_by()` and `count()` to produce a summary of the factor column:

```{r}
stoves %>% 
group_by(type) %>%
count()
```

---

# Ordering Factors

Factors may not be a useful order. There are a number of ways to a factor:

## by variable

```{r}
 stove_exposure <- stoves %>%
                   left_join(data_frame(type = factor(levels),
                                        pm25 = c(10, 500, 35, 100, 250, 0)),
                             by = "type")
```

We can use `fct_reorder()` to reorder the levels in a factor by a different variable. For example, let's reorder `type` by the $PM_{2.5}$ exposure level:

```{r}
stove_exposure <- stove_exposure %>%
                  mutate(type =  fct_reorder(type, pm25))

levels(stove_exposure$type)
```

## by frequency

We can use `fct_infreq()` to reorder the levels in a factor by the frequency that they occur. For example, let's reorder `type`:

```{r}
stove_exposure <- stove_exposure %>%
                  mutate(type =  fct_infreq(type))

levels(stove_exposure$type)
```

## by first

We can use `fct_inorder()` to reorder the levels in a factor by order in which they appear in the vector:

```{r, echo = FALSE}
stove_exposure$type[1:10]
```


```{r}
fct_inorder(stove_exposure$type) %>%
levels()
```

---

# Manipulating Levels

## move

The `fct_rev()` functions reverses the order of a factor:

```{r}
levels(stove_exposure$type)
```

```{r}
levels(fct_rev(stove_exposure$type))
```

We can use the `fct_relevel()` function to move the position of factors.

* move `control` to the first position:

```{r}
levels(fct_relevel(stove_exposure$type, "control"))
```

* move `three_stone` to the end position:

```{r}
fct_relevel(stove_exposure$type, "three_stone", after = Inf) %>%
levels()
```

## change

To change levels we can use the `fct_recode()` function:

```{r}
fct_recode(stove_exposure$type,
           "Control" = "control",
           "LPG" = "lpg",
           "Gasifier" = "gasifier",
           "Fan Rocket" = "fan_rocket",
           "Rocket Elbow" = "rocket_elbow",
           "Three Stone Fire" = "three_stone") %>%
levels()
```

## combine

```{r}
fct_collapse(stove_exposure$type,
             wood = c("gasifier", "fan_rocket",
                      "rocket_elbow", "three_stone")) %>%
levels()

```

```{r}
fct_lump(stove_exposure$type, n = 3) %>%
levels()
```

We can use the `fct_other()` to combine levels. The `keep` argument allows us to specify the levels to preserve:


```{r}
fct_other(stove_exposure$type,
          keep = c("control", "three_stone"),
          other_level = "other_stove") %>%
levels()
```

The `drop` argument allows us to specify the levels to remove:

```{r}
fct_other(stove_exposure$type,
          drop = c("control", "three_stone"),
          other_level = "other_stove") %>%
levels()
```

---

